\documentclass[11pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[margin=2.5cm]{geometry}
\usepackage{graphicx} 
\usepackage{listings} 
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{caption}
\usepackage[hidelinks]{hyperref}

\definecolor{codegray}{gray}{0.95}
\definecolor{commentgreen}{rgb}{0.0,0.5,0.0}
\definecolor{keywordblue}{rgb}{0.0,0.0,0.7}
\definecolor{stringred}{rgb}{0.6,0.0,0.0}

\lstdefinestyle{cpp}{
    language=C++,
    backgroundcolor=\color{codegray},
    basicstyle=\ttfamily\footnotesize,
    keywordstyle=\color{keywordblue},
    commentstyle=\color{commentgreen},
    stringstyle=\color{stringred},
    showstringspaces=false,
    breaklines=true,
    frame=single,
    numbers=left,
    numberstyle=\tiny,
    captionpos=b,
    tabsize=2
}
\lstset{style=cpp}

\title{A Particle Detector Simulation in C++}
\author{Roman le Bon \\ University of Manchester \\ PHYS30762 Object-Oriented Programming in C++}
\date{\today}

\begin{document}
\maketitle
% ABSTRACT  -- ~100-150 words. Write LAST.
% Rubric: states what was built, which OOP concepts were used, one result.
% Must NOT just restate the brief. Must show reasoning, not be purely descriptive.
\begin{abstract}
% TODO
\end{abstract}

% 1. INTRODUCTION  -- ~20% of length (~500 words)
% Rubric: aim and structure understandable by a reader who does NOT know
% the code. Not too complex, not identical to the abstract. Reasoning
% must be evident here too, not just description.

\section{Introduction}
This project implements a simplified particle detector simulation in C++, inspired by the Compact Muon Solenoid (CMS) experiment at the Large Hadron Collider (LHC). CMS is a general-purpose detector at one of the LHC collision points, designed to identify stable particles produced in collusions and measure their momenta and energies so that the collision can be reconstructed for further analysis \cite{cmsdetector}. The aim of this project was not to reproduce the full complexity of CMS, but instead to build a smaller object-orientated model that captures the same basic idea. That is, that particles pass through several detector layers and produce different patterns of response. 

The physical model used in the program is based on a CMS-like detector made from sub-detector components. In the simulation, charged particles can be measured by a tracker, particles depositing energy can be measure by a calorimeter, and muons can pass through to muon chambers. Neutral weakly interacting particles (like neutrinos) are not directly detected in this simplified modeel, instead contributing to missing  energy. Therefore, the program compares each particle's true energy with the energy measured by each detector component, then uses the pattern of measurements to then infer the likely particle type.

The coding aim was to represent this physical situation using an object-orientated C++ design. The code is built around two min class hierarchies - one for particles and one for detector components. The particle hierarchy stores common physcial quantities such as four-momentum, spin and rest mass, whilst derived classes represent more specific particle types such as electrons, muons, taus, neutrinos, photons and hadrons. The detector hierarchy represent sub-detectors such the tracker, calorimeter and muon chamber - each of which implements its own detection rule. 

The above structure was chosen because the problem naturally involves specialised behaviour and shared interfaces and specialised behaviour. All particles share common physical properties, but different particle types need different additional data (such as lepton number, charge, lifetime and-or flavour). Similarly, all suub-detectors can attempt to measure a particle, but each sub-detector responds differntly. Thus, abstract base classes and virtual functions allow the main detector code to treat particles and sub detectors more generically, while still calling the correct derived class behaviour during run time. The final program reads particles from an input txt file, then constructing a CMS-like detector from sub-detector components, passes each particle through a detector, then printing the true and measured energies, inferring the liekly particle type from the detector signals and calculating missing energy from invisible particles. Additional functionality includes a realsitic calorimeter mode with Gaussian (normal) noise and exception handling for a particle input that is invalid. 

The next part of the paper will look
% 2. CODE DESIGN AND IMPLEMENTATION  -- ~50% of length (~1250 words)
% Rubric: emphasise advanced features; short, professionally-formatted
% code snippets; class diagram; reasoning behind EVERY advanced-feature
% decision (2 marks rely on this); reflection tied to actual classes.
% This is where the working notes get reshaped into argued prose.

\section{Code Design and Implementation}

\subsection{Class Hierarchies}
% TODO - the two hierarchies (particles, sub-detectors), why abstract
% base classes. Inheritance diagram goes here.

% Inheritance diagram - proper UML, made in draw.io, exported as PDF/PNG.

\subsection{Key Design Decisions}
% TODO - FourMomentum as std::array, perfect/realistic detector split,
% inference logic, the missing-energy max-reading fix. Each: decision +
% why + what the alternative was.

\subsection{Advanced C++ Features}
% TODO - smart pointers (unique_ptr vs shared_ptr - the deliberate
% contrast), <random> for noise, exceptions, static, std::map, file I/O.
% Each with a SHORT snippet and specific reasoning.

% 3. RESULTS  -- ~15% of length (~375 words)
% Rubric: illustrate how the code is used, including input AND output.

\section{Results}
% TODO - describe the input file particles.txt.
% TODO - include a short excerpt of particles.txt showing the format.
% TODO - show representative output from one or two particles.
% TODO - include detector measurements from Tracker, Calorimeter and MuonChamber.
% TODO - include one inference example.
% TODO - include missing energy result, e.g. 15 MeV if this is still the final
% output.
% TODO - explain briefly why the output validates the design.

% 4. DISCUSSION AND CONCLUSIONS  -- ~10% of length (~250 words)
% Rubric: concrete, implementable improvements. Do not just say "add more
% features". The challenge mark wants explicit discussion of extending YOUR
% OWN classes and design.

\section{Discussion and Conclusions}
% TODO - summarise whether the program met the aim.
% TODO - discuss limitations:
% no ECAL/HCAL split, so calorimeter response is simplified;
% tau decay products are not simulated;
% realistic calorimeter noise is simple Gaussian noise;
% inference logic is rule-based and could be generalised.
% TODO - discuss concrete extensions:
% split Calorimeter into ECAL and HCAL;
% add tau decay products;
% add more detailed detector resolution;
% replace detector-name checks in inference with a more general interface.
% TODO - explicitly mention how the class hierarchy makes these extensions
% possible.

% REFERENCES  -- not counted in word limit. Use one consistent style.
% Add references once the prose is written.
% CMS detector overview liekly reference

\begin{thebibliography}{9}

\bibitem{cmsdetector}
CMS Collaboration,
\textit{Detector},
CERN, CMS Experiment.
Available at: \url{https://cms.cern/detector}
[Accessed May 2026].

\end{thebibliography}

\end{document}
